---
title: "Lab 3 Marijn"
author: "Marijn Jaarsma"
date: "2023-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries, echo=FALSE, warning=FALSE, message=FALSE}
library(geosphere)

```

# 1. Kernel Methods
## Setup
```{r Q1_Setup}
# Read in data and merge
df_stations <- read.csv("stations.csv", fileEncoding="latin1")
df_temps <- read.csv("temps50k.csv")
df_st <- merge(df_stations, df_temps, by="station_number")

# Define hyper parameters
h_distance <- 4430000
h_date <- 10000
h_time <- 8

# Prediction point
a <- 58.4274
b <- 14.826
# a <- 55.386246
# b <- 13.056955
date <- "2013-06-04"
times <- c("04:00:00", "06:00:00", "08:00:00", "10:00:00", "12:00:00", "14:00:00", "16:00:00", "18:00:00", "20:00:00", "22:00:00", "24:00:00")

```

## Kernel Computation (Sum of Three Gaussian Kernels)
```{r Q1_Code}
# Remove rows with observations posterior to wanted time
data_dates <- strptime(df_st[, "date"], "%Y-%m-%d", tz="UTC")
wanted_date <- strptime(date, "%Y-%m-%d", tz="UTC")
df_st_filt <- df_st[which(data_dates < wanted_date),]

# Distance kernel
k_dist <- exp(-(distHaversine(df_st_filt[, c("longitude", "latitude")], c(a, b))^2) / (2 * h_distance^2))  

plot(distHaversine(df_st_filt[, c("longitude", "latitude")], c(a, b)),
     k_dist,
     xlab="Distance (m)", ylab="Kernel Value (Gaussian)")
title(paste0("Distance Kernel (h=", h_distance, ")"))

### FIND BEST h_distance ###
# v <- vector()
# for (i in seq(1000000, 7000000, by=10000)) {
#   k_dist <- exp(-(distHaversine(df_st_filt[, c("longitude", "latitude")], c(a, b))^2) / (2 * i^2))  
#   v <- append(v, max(k_dist) - min(k_dist))
# }
# 
# plot(seq(1000000, 7000000, by=10000), v)

# seq(1000000, 7000000, by=10000)[which(v == max(v))]
############################

# Date kernel
v_date_differences <- as.numeric(as.Date(date, "%Y-%m-%d") - as.Date(df_st_filt[, "date"], "%Y-%m-%d"))
k_date <- exp(-(v_date_differences^2) / (2 * h_date^2))

plot(v_date_differences, k_date,
     xlab="Distance (days)", ylab="Kernel Value (Gaussian)")
title(paste0("Date Kernel (h=", h_date, ")"))

# Hour kernel
# https://stackoverflow.com/questions/53818163/calculate-difference-between-hours-not-considering-date
data_hour <- as.numeric(substr(df_st_filt[, "time"], 1, 2))
wanted_hours <- matrix(rep(sapply(times, function(x) {as.numeric(substr(x, 1, 2))}), nrow(df_st_filt)), ncol=length(times), byrow=TRUE)
v_time_differences <- abs(sweep(wanted_hours, 1, data_hour))
k_time <- exp(-(v_time_differences^2) / (2 * h_time^2))

plot(v_time_differences[, 1], k_time[, 1],
     xlab="Distance (hours)", ylab="Kernel Value (Gaussian)")
title(paste0("Time Kernel (h=", h_time, ")"))

# Combined kernel summed
k_comb_sum <- sweep(k_time, 1, k_dist + k_date, FUN="+")

# Non-parametric kernel regression
temp_sum <- colSums(k_comb_sum * df_st_filt$air_temperature) / colSums(k_comb_sum)
plot(temp_sum, type="o", xaxt="n", xlab="Time", ylab="Predicted Temperature")
axis(1, at=1:length(temp_sum), labels=times)
title(paste0("Summed Kernel Predictions\n", date))

```

## Kernel Computation (Product of Three Gaussian Kernels)
```{r Q1_Mult_Kernel}
# Combined kernel product
k_comb_prod <- k_dist * k_date * k_time

# Non-parametric kernel regression
temp_prod <- colSums(k_comb_prod * df_st_filt$air_temperature) / colSums(k_comb_prod)
plot(temp_prod, type="o", xaxt="n", xlab="Time", ylab="Predicted Temperature")
axis(1, at=1:length(temp_prod), labels=times)
title(paste0("Product Kernel Predictions\n", date))

```

## Kernel Comparison
The product kernel produces slightly higher values than the summed kernel. This is due to the division in the computation of the predictions. Multiplying the kernels gives smaller values than summing would, as the kernel values are all between 0 and 1 due to the fact that a Gaussian kernel function was used. However, when dividing by a smaller value this produces slightly higher values than the summed kernel produces. 

The model is not very flexible to different inputs; a different location or time of year gives almost the same result. Perhaps different weights should be attached to the different kernels. The hour kernel is always the same, and has equal weight to the other two kernels. Additionally, the distance kernel is only on a range between about 0.32 and 0.42 rather than 0 and 1 as the distance values in this kernel are so large. A big difference in distance may have a big impact on the predicted temperature, but this doesn't reflect from the kernel value. A linear combination between the kernels, where the weights may be tuned may be a better option.